% Seleção de modelos de crescimento para descrição da área foliar do linho 

```{r global_options, include = FALSE}
knitr::opts_chunk$set(cache = FALSE,
                      comment = "##",
                      collapse = TRUE,
                      warning = FALSE,
                      message = FALSE)

```


# Pacotes

To reproduce the examples of this material, the R packages the following packages are needed.

```{r warning=FALSE, message=FALSE}
library(rio)
library(tidyverse)
library(car)
library(nlstools)
library(lmtest)
library(MASS)
library(nlstools)

my_theme <- 
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(), #remove minor gridlines
    axis.text = element_text(colour = "black"),
    axis.title = element_text(color = "black")
  )
```


# Raw data
```{r}
df_nlm <-  
  import("data/df_models.xlsx") |> 
  group_by(das) |> 
  summarise(af_planta = mean(af_planta))
```


# Logistic
## Fitted
```{r}
logi <- af_planta ~ b1/(1 + exp(b2 - b3 * das))
start_logi <- c(b1 = 120,
                b2 = 3,
                b3 = 0.05)
mod_logi <- 
  nls(logi,
      data = df_nlm,
      start = start_logi)
summary(mod_logi)
```

## Model assumptions
```{r}
shap_logi <- shapiro.test(residuals(mod_logi))
# obtaining the gradient matrix
grad_logi <- attr(mod_logi$m$fitted(),"gradient")
# gradient matrix after removing intercept
m0_logi <- lm(af_planta ~ -1 + grad_logi, data = df_nlm)
# Durbin-Watson test for independence of residuals
(DW_logi <- durbinWatsonTest(m0_logi))
# Breusch-Pagan' test for heteroscedasticity of residuals
(BP_logi <- bptest(m0_logi))
```

## Nonlinearity estimates
```{r}
mcurv_logi <- 
  deriv3( ~ b1/(1 + exp(b2 - b3 * das)),
          c("b1", "b2","b3"),
          function(b1, b2, b3, das) NULL)

logi_nl <- 
  nls(af_planta ~ mcurv_logi(b1, b2, b3, das),
      data = df_nlm,
      start = start_logi)
(nl_logi <- rms.curv(logi_nl))
d <- nlsContourRSS(mod_logi)
par(mfrow=c(1,3))
plot(d, nlev = 5, nrow = 2)

```

## Confidence intervals
```{r}
conf_logi <- nlsBoot(mod_logi)
conf_logi$bootCI
```




# Gompertz
## Fitted
```{r}
gomp <- af_planta ~ b1 * exp(-b2*exp(-b3*das))
start_gomp <- c(b1 = 120,
                b2 = 40,
                b3 = 0.05)
mod_gomp <- 
  nls(gomp,
      data = df_nlm,
      start = start_gomp)
summary(mod_gomp)
```

## Model assumptions
```{r}
(shap_gomp <- shapiro.test(residuals(mod_gomp)))
# obtaining the gradient matrix
grad_gomp <- attr(mod_gomp$m$fitted(),"gradient")
# gradient matrix after removing intercept
m0_gomp <- lm(af_planta ~ -1 + grad_gomp, data = df_nlm)
# Durbin-Watson test for independence of residuals
(DW_gom <- durbinWatsonTest(m0_logi))
# Breusch-Pagan' test for heteroscedasticity of residuals
(BP_gom <- bptest(m0_logi))
```


## Nonlinearity estimates
```{r}
mcurv_gomp <- 
  deriv3( ~ b1 * exp(-b2*exp(-b3*das)),
          c("b1", "b2","b3"),
          function(b1, b2, b3, das) NULL)

gomp_nl <- 
  nls(af_planta ~ mcurv_gomp(b1, b2, b3, das),
      data = df_nlm,
      start = start_gomp)
(nl_gom <- rms.curv(gomp_nl))
d <- nlsContourRSS(gomp_nl)
plot(d, nlev = 10)

```

## Confidence intervals
```{r}
conf_gom <- nlsBoot(mod_gomp)
conf_gom$bootCI
```









# Von bertallanfy
## Fitted
```{r}
vbl <- af_planta ~ b1 *(1-b2*exp(-b3*das))^3
start_vbl <- c(b1 = 50,
               b2 = 100,
               b3 = 0.01)
mod_vbl <- 
  nls(vbl,
      data = df_nlm,
      start = start_vbl)
summary(mod_vbl)
```

## Model assumptions
```{r}
(shap_vbl <- shapiro.test(residuals(mod_vbl)))
# obtaining the gradient matrix
grad_vbl <- attr(mod_vbl$m$fitted(),"gradient")
# gradient matrix after removing intercept
m0_vbl <- lm(af_planta ~ -1 + grad_vbl, data = df_nlm)
# Durbin-Watson test for independence of residuals
(DW_vbl <- durbinWatsonTest(m0_vbl))
# Breusch-Pagan' test for heteroscedasticity of residuals
(BP_vbl <- bptest(m0_vbl))
```


## Nonlinearity estimates
```{r}
mcurv_vbl <- 
  deriv3( ~  b1 *(1-b2*exp(-b3*das))^3,
          c("b1", "b2","b3"),
          function(b1, b2, b3, das) NULL)

vbl_nl <- 
  nls(af_planta ~ mcurv_vbl(b1, b2, b3, das),
      data = df_nlm,
      start = start_vbl)
(nl_vbl <- rms.curv(vbl_nl))
d <- nlsContourRSS(vbl_nl)
plot(d, nlev = 10)

```

## Confidence intervals
```{r}
conf_vbl <- nlsBoot(mod_vbl)
conf_vbl$bootCI
```


# Fitted model
```{r}
ggplot(df_nlm, aes(das, af_planta)) +
  geom_point(size = 3) +
  geom_smooth(method = "nls",
              method.args = list(formula =  y ~ b1/(1 + exp(b2 - b3 * x)),
                                 start = start_logi),
              aes(color = "Logístico"),
              se = FALSE) +
  geom_smooth(method = "nls",
              method.args = list(formula =  y ~ b1 * exp(-b2 * exp(-b3 * x)),
                                 start = start_gomp),
              aes(color = "Gompertz"),
              se = FALSE) +
  geom_smooth(method = "nls",
              method.args = list(formula =  y ~  b1 *(1-b2*exp(-b3*x))^3,
                                 start = start_vbl),
              aes(color = "von Bertalanffy"),
              se = FALSE) +
  # scale_y_continuous(breaks = seq(0, 150, by = 25)) +
  labs(x = "Dias após a semeadura",
       y = expression(Área~foliar~(cm^2~planta^{-1}))) +
  scale_x_continuous(breaks = seq(20, 120, by = 10)) +
  my_theme +
  theme(legend.position = c(.2,.8),
        legend.box.background = element_blank(),
        legend.background = element_blank(),
        legend.key = element_blank(),
        legend.title = element_blank())

```






# Tables
```{r}
observed <- df_nlm$af_planta
gof_logi <- hydroGOF::gof(observed, fitted(mod_logi), digits = 5) |> t()
gof_gom <- hydroGOF::gof(observed, fitted(mod_gomp), digits = 5) |> t()
gof_vbl <- hydroGOF::gof(observed, fitted(mod_vbl), digits = 5) |> t()
gofs <- rbind(gof_logi, gof_gom, gof_vbl) |> as.data.frame()


parameters <- 
  cbind(data.frame(model = c("Logístico", "Gompertz", "von Bertalanffy")),
        rbind(coef(mod_logi), coef(mod_gomp), coef(mod_vbl)),
        shapi = c(shap_logi$p.value, shap_gomp$p.value, shap_vbl$p.value),
        BP = c(BP_logi$p.value, BP_gom$p.value, BP_vbl$p.value),
        DW = c(DW_logi$p, DW_gom$p, DW_vbl$p),
        nlpar = c(nl_logi$pe, nl_gom$pe, nl_vbl$pe),
        nlint = c(nl_logi$ic, nl_gom$ic, nl_vbl$ic),
        AIC = AIC(mod_logi, mod_gomp, mod_vbl)$AIC,
        R2 = gofs$R2,
        RMSE = gofs$RMSE)
export(parameters, "data/parameters_models.xlsx")
```



## First order derivate
```{r}
# primeira derivada
D(expression(b1/(1 + exp(b2 - b3 * das))), "das")

dy <- function(x,b1,b2,b3){
  b1 * (exp(b2 - b3 * x) * b3)/(1 + exp(b2 - b3 * x))^2
}

parameters <- 
  parameters |> 
  mutate(xpi = b2 / b3,
         ypi = dy(xpi, b1, b2, b3)) |> 
  as.data.frame()

plot_pi <-
  ggplot() + 
  stat_function(fun = dy,
                size = 1,
                xlim = c(20, 120),
                args = c(b1 = parameters[[1, 2]],
                         b2 = parameters[[1, 3]],
                         b3 = parameters[[1, 4]])) +
  geom_point(aes(67.30368, 4.995532),
             size = 3,
             show.legend = FALSE) +
  scale_x_continuous(breaks = seq(20, 120, by = 10)) +
  labs(x = "Dias após a semeadura",
       y = expression(Emissão~de~área~foliar~(cm^2~planta^{-1}~dia^{-1}))) +
  my_theme

  
  
  
  
  # segunda derivada
D(expression(b1 * (exp(b2 - b3 * x) * b3)/(1 + exp(b2 - b3 * x))^2), "x")

d2y <- function(x,b1,b2,b3){
  -(b1 * (exp(b2 - b3 * x) * b3 * b3)/(1 + exp(b2 - b3 * x))^2 - 
      b1 * (exp(b2 - b3 * x) * b3) * (2 * (exp(b2 - b3 * x) * b3 * 
                                             (1 + exp(b2 - b3 * x))))/((1 + exp(b2 - b3 * x))^2)^2)
}

parameters <- 
  parameters |> 
  mutate(xmap = (b2 - 1.3170)/b3,
         xmdp = (b2 + 1.3170)/b3,
         ymap = d2y(xmap, b1, b2, b3),
         ymdp = d2y(xmdp, b1, b2, b3),
  )

# plota somente o logístico

df_acel <-
  ggplot() + 
  geom_hline(yintercept = 0) +
  stat_function(fun = d2y,
                size = 1,
                xlim = c(20, 120),
                args = c(b1 = parameters[[1, 2]],
                         b2 = parameters[[1, 3]],
                         b3 = parameters[[1, 4]])) +
  geom_point(aes(xmap, ymap),
             data = parameters[1, ],
             size = 3,
             shape = 19,
             show.legend = FALSE) +
  geom_point(aes(xmdp, ymdp),
             data = parameters[1, ],
             size = 3,
             shape = 17,
             show.legend = FALSE) +
  geom_point(aes(xpi, 0),
             data = parameters[1, ],
             size = 3,
             shape = 15,
             show.legend = FALSE) +
  scale_x_continuous(breaks = seq(20, 120, by = 10)) +
  labs(x = "Dias após a semeadura",
       y = "Aceleração da emissão de área foliar") +
  my_theme

metan::arrange_ggplot(plot_pi, df_acel,
                      tag_levels = "a")

```




# Section info
```{r}
sessionInfo()
```

